Hydras \& Co.

### 2023-06-05 doc/chapter-fol.tex
Remove Export FolNotations
Removed  useless Exports
redaction of chapter 12 (in progress)

\begin{remark}
The type \texttt{Formula} (over $L$) has only five constructors,
respectively bound to term equalities $t_1=t_2$, atomic formulas of the form $(r\;t_1\;\dots\;t_n)$, implications $(F \arrow G)$, negations $({\sim}F)$, and universal quantifications $(\forall\,i, F)$.

In~\cite{Goedel}, no \emph{constructors} are associated with
disjunction, conjonction, logical equivalence and existential quantifier. These constructs are formalized through \emph{definitions}\footnote{Please keep in mind that we are considering a classical logic.}.

@@

Let us give a few examples of first-order formulas over $L$.
\label{fol:examplesf1f2f3}
\begin{enumerate}
\item $R\;a\;b$
\item $\forall v_0\; v_1, R\;v_0\;v_1 \arrow R\;v_1\;v_0$
\item $ \forall v_0, v_0=a \vee \exists\;v_1,\, v_0= f(v_1)$
\item $(\forall v_1, v_0 = v_1) \vee \exists\,v_0\,v_1, v_0 \not= v_1$
\end{enumerate}

@@

  In~\cite{Goedel}, there is no De Bruijn encoding of bound variables (see also~\cite{OConnor05}). The following example shows two formulas which share the same structure, are logically equivalent, but are not Leibniz equal.

@@

In the formula \texttt{F4} \vpageref[above]{sect:folFormExamples}, the  term \texttt{(var 0)} occurs freely and also inside the scope of a quantifier.

@@

 We use \coq's \texttt{Notation} features to print and parse terms and formulas  in a more readable form.
To this purpose, we build \texttt{fol\_scope}, a  notation scope
where the main connectives and quantifiers get a syntax close to \coq's.  Additionnally, a term of the form \texttt{(var $i$)} is
just written \texttt{v\#$i$}.

@@

@@

\subsection{On variables}

\begin{todo}
  Difference from HOAS. Free and bound occurrences.
\end{todo}

\subsection{Decidability of equality}
Let $L$ be a language, and let us assume that equality 
of function and relation symbols of $L$ are decidable.
Under this assumption, equality of terms and formulas over $L$ is decidable too.

@@

\begin{todo}
  Should we talk about the decomposition lemmas
\texttt{nilTerms} and \texttt{consTerms} ?
\end{todo}

@@

\subsection{Induction principles on formula depth}

\inputsnippets{FolExamples/substNotStructural}

The purpose of this section is to present ways of proving 
properties of any formula $a$ (or defining functions over the type of formulas).

\subsubsection{Depth of a formula}
As explained in~\cite{OConnor05}, some functions on the \texttt{Formula} type cannot be defined by structural recursion in 
O'Connors formalization. For instance, the substitution of a term to a free variable (because of quantifiers). 
In such cases, we may define functions by well-founded recursion. In this section, we consider the well-founded strict-order associated with a measure: the \emph{depth} of a formula.

\begin{todo}
Forward link to its application.
\end{todo}

The function \texttt{depth} computes the \emph{depth} of the 
 skeleton of any formula.

@@

The function \texttt{depth} allows us to define a strict order on 
\texttt{Formula $L$}.

@@

\section{Languages for Arithmetic}

@@

\begin{todo}
Fix the issue about the way formulas are printed (mix of different scopes).
\end{todo}
\inputsnippets{LNN_Examples/uglyF0a}

>>>>

\begin{remark}
The type \texttt{Formula} (over $L$) has only five constructors,
respectively associated with term equalities $t_1=t_2$, atomic formulas of the form $(r\;t_1\;\dots\;t_n)$, implications $(F \arrow G)$, negations $({\sim}F)$, and universal quantifications $(\forall\,i, F)$.

In~\cite{Goedel}, no \emph{constructors} of type \texttt{(Formula L)} are associated with
disjunction, conjonction, logical equivalence and existential quantifier. These constructs are formalized through \emph{definitions}\footnote{Please keep in mind that we are considering a classical logic.}.

@@

Let us give a few examples of first-order formulas over $L$.
\label{fol:examplesf1f2f3}
\begin{description}
\item[F1] $R\;a\;b$
\item[F2] $\forall v_0\; v_1, R\;v_0\;v_1 \arrow R\;v_1\;v_0$
\item[F3] $ \forall v_0, v_0=a \vee \exists\;v_1,\, v_0= f(v_1)$
\item[F4] $(\forall v_1, v_0 = v_1) \vee \exists\,v_0\,v_1, v_0 \not= v_1$
\item[F5] $v_0 =a \vee v_0 = f(v_1)$
 \item[F6] $\forall v_0,\,\exists v_1, v_0= f(v_1)\wedge v_0\not= v_1$
\end{description}

Let us now define these formulas as terms of type
\texttt{(Formula L)}.

@@

In~\cite{Goedel}, there is no De Bruijn encoding of bound variables (see also~\cite{OConnor05}).

For instance, the the  term \texttt{(var 0)}
occurs freely and also inside the scope of a quantifier in the
formula \texttt{F4} \vpageref[above]{sect:folFormExamples}.

The following example shows two formulas which share the same structure, are logically equivalent, but are not Leibniz equal.

@@

I

@@

 We use \coq's \texttt{Notation} features to print and parse terms and formulas  in a more readable form.
To this purpose, we build \texttt{fol\_scope}, a  notation scope
where the main connectives and quantifiers get a syntax close to \coq's.  Additionnally, a term of the form \texttt{(@var \_ $i$)} is
just printed and parsed  \texttt{v\#$i$}.

@@

\inputsnippets{FolExamples/toyNotationForm4}

@@

\subsection{Decidability of equality}
% Deciding whether two formulas are equal is a part of the verification of a proof (for instance in the implementation of the \emph{assumption} rule).

Let $L$ be a language, and let us assume that equality 
of function and relation symbols of $L$ are decidable.
Under this assumption, equality of terms and formulas over $L$ is decidable too.

@@

@@

\begin{remark}
Please note that \texttt{term\_dec}, \texttt{terms\_dec} and
\texttt{formula\_dec} are \emph{opaque}.

The function \texttt{formula\_dec} is mainly used in
\href{../theories/html/hydras.Ackermann.PA.html}{Ackermann.PA}, in order to check whether a given formula belongs to the axioms of Peano arithmetic.
\begin{todo}
  Look for the use of \texttt{open}  (in codePA)
\end{todo}

\subsection{Variables and substitutions}

Since free and bound occurrences of a variable $i$ are represented the same way, much care should be taken in programming the substitution of a term to a variable in order to avoid \emph{variable capture}.

For instance let $F$ be the formula 
$\exists v_2, v_1 \not= f(v_2)$. The result of the substitution 
of the term $f(v_2)$ to the (free) occurrences of $v_1$ in $F$
 \emph{is not} $\exists v_2, f(v_2) \not= f(v_2)$, but, for instance $\exists v_3, f(v_2) \not= f(v_3)$, where $v_3$ is a ``fresh'' variable.

\inputsnippets{FolExamples/substExample1}
\inputsnippets{FolExamples/substExample2}

From this little example, we can guess that the function
\texttt{substF} which is assumed to compute the substitution in a formula $F$ the substitution of a term $t$ to the free occurrences of a variable $v$ will not be implemented as a direct structural recursive function. For instance, the formula $G$, obtained through a variable renaming, is not a strict subterm of $F$.

To fix this issue, we can associate  a measure (\emph{e.g.} a natural number) to each formula, such that the renaming of a strict subformula $F$ is strictly smaller that $F$.

\subsubsection{Depth of a formula}

The function \texttt{depth} computes the \emph{depth} of the 
 skeleton of any formula, and allows us to define a well-founded strict order on 
\texttt{Formula $L$}.

@@

@@

\begin{todo}
Make a chapter!
\end{todo}
\section{Languages for Arithmetic}

@@
